---
title: "Binary ODA: Gully Erosion Adjustment and Motivation"
author: "odacore"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binary ODA: Gully Erosion Adjustment and Motivation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Research question

A study of gully erosion in rural southeast Nigeria documented the type of
adjustment made by community members in response to erosion hazards, and asked
whether adjustment type could distinguish individually-motivated from
community-motivated responses.^[Okuh D, Osumgborogwu IE (2019). Adjustments to
hazards of gully erosion in rural southeast Nigeria: A case of Amacha
communities. *Applied Ecology and Environmental Sciences*, 7, 11-20.]

Four adjustment types were recorded: (1) Use of Ridges, (2) Shifting
Habitation, (3) Relocation, and (4) Intensified Cultivation. Because Ridges
and Shifting Habitation are actions that can be taken individually or
collectively, whereas Relocation and Intensified Cultivation require organised
community effort, the hypothesis is that Ridges and Shifting predict Community
motivation and Relocation and Intensification predict Individual motivation.
Optimal Data Analysis (UniODA) tests whether adjustment type discriminates
motivation and quantifies the strength of the association.

## Data

Motivation (0 = Individual, 1 = Community) is the class variable; adjustment
type (1-4) is the attribute. Published cell frequencies are reconstructed
directly into observation-level vectors  -  no external data file is required.

```{r data}
library(odacore)

# Cross-classification: rows = adjustment type, cols = motivation.
#                    Indiv (0)  Comm (1)   total
#  Ridges       (1)      85       173       258
#  Shifting     (2)      65       170       235
#  Relocation   (3)     172        10       182
#  Intensified  (4)      45         0        45
#                       367       353       720

motivation  <- c(rep(0L,  85), rep(1L, 173),   # adjustment = 1
                 rep(0L,  65), rep(1L, 170),   # adjustment = 2
                 rep(0L, 172), rep(1L,  10),   # adjustment = 3
                 rep(0L,  45), rep(1L,   0))   # adjustment = 4
adjustment  <- c(rep(1L, 258), rep(2L, 235),
                 rep(3L, 182), rep(4L,  45))

table(adjustment, motivation,
      dnn = c("Adjustment (1=Ridges,2=Shifting,3=Relocation,4=Intensified)",
              "Motivation (0=Individual, 1=Community)"))
```

## Fit the ODA model

Adjustment type is a four-category nominal variable. Because the directional
hypothesis specifies *a priori* that Ridges and Shifting predict Community
motivation and Relocation and Intensified Cultivation predict Individual
motivation, a fixed-partition `direction_map` is supplied. ODA evaluates only
this mapping; no partition search is performed. This is MPE Chapter 4
categorical fixed-partition DIRECTIONAL (Phase 6C). Leave-one-out (LOO)
jackknife validity analysis is included; because the mapping is fixed, LOO
predictions are invariant across folds (trivially stable by construction).

```{r fit-canonical, eval=FALSE}
# Canonical reference run (mc_iter = 25000L; not evaluated in CRAN vignette)
fit <- oda_fit(
  x             = adjustment,
  y             = motivation,
  attr_type     = "categorical",
  direction_map = c("1" = 1L, "2" = 1L, "3" = 0L, "4" = 0L),
  mc_iter       = 25000L,
  loo           = "on"
)
```

```{r fit}
# CRAN-safe run: mc_iter = 500L for vignette rendering speed.
# Training rule, ESS, and confusion matrix are identical to the canonical run.
fit <- oda_fit(
  x             = adjustment,
  y             = motivation,
  attr_type     = "categorical",
  direction_map = c("1" = 1L, "2" = 1L, "3" = 0L, "4" = 0L),
  mc_iter       = 500L,
  mc_seed       = 42L,
  loo           = "on"
)
```

## Rule and confusion matrix

```{r print-fit}
print(fit)
```

ODA identified a binary partition of the four adjustment categories:

- If adjustment in {1, 2} (Ridges or Shifting) -> predict motivation = Community (1)
- If adjustment in {3, 4} (Relocation or Intensified) -> predict motivation = Individual (0)

This rule is consistent with the directional hypothesis.

```{r confusion}
# Confusion matrix: actual motivation (rows) x predicted motivation (cols)
conf_mat <- matrix(
  c(fit$confusion$TN, fit$confusion$FP,
    fit$confusion$FN, fit$confusion$TP),
  nrow = 2L, byrow = TRUE,
  dimnames = list(Actual    = c("Indiv(0)", "Comm(1)"),
                  Predicted = c("Indiv(0)", "Comm(1)"))
)
print(conf_mat)
```

## ESS / PAC / PV interpretation

```{r metrics}
summary(fit)
```

```{r pv}
# Predictive value: accuracy when the model makes a prediction into each class
pv_indiv <- fit$confusion$TN / (fit$confusion$TN + fit$confusion$FN)
pv_comm  <- fit$confusion$TP / (fit$confusion$TP + fit$confusion$FP)
cat("PV Individual (0):", round(pv_indiv * 100, 1), "%\n")
cat("PV Community  (1):", round(pv_comm  * 100, 1), "%\n")
```

- **PAC (sensitivity per class):** 59.1% for Individual members, 97.2% for
  Community members. Because 50% correct per class is expected by chance, the
  model classifies Community-motivated adjustments nearly twice as well as chance
  while Individual classification exceeds chance by only 18%.
- **ESS = 56.30%** indicates a relatively strong effect.^[Yarnold, P.R., &
  Soltysik, R.C. (2005). *Optimal Data Analysis: A Guidebook with Software for
  Windows.* Washington, D.C.: APA Books.] The pronounced asymmetry (59% vs. 97%)
  reflects near-uniform community uptake of Ridges and Shifting adjustments.
- **PV:** When the model predicts Individual motivation it is correct ~95.6% of
  the time; when it predicts Community motivation ~69.6%.

## Monte Carlo and LOO validity

The MC p-value and LOO result are shown in the `summary` output above.

- **LOO stability:** Because the prediction rule is fixed *a priori* by
  `direction_map`, every LOO fold applies the same mapping to the same x
  values. LOO ESS equals the training ESS (56.30%) by construction  - 
  trivially stable.
- **LOO Fisher exact p < .001:** Statistical significance confirmed in holdout
  analysis.
- **MC p-value:** The fixed partition is evaluated on each permuted y vector,
  yielding a directional Fisher-randomization p-value. The `p(MC)` shown
  is expected to be small (MegaODA reports p = 0.000000); the 500-iteration
  CRAN run may show p = 0.0. Use `mc_iter = 25000L` for publication results.

## Notes on reproducibility

**Fixture parity.** The training rule, confusion matrix, and ESS are verified
against MegaODA.exe output in the package test suite
(`tests/testthat/test-fixture-vignettes.R`, Example 2).

**MC p-value calibration.** The MC p shown here may differ from MegaODA's
reported p = 0.000000 (exact zero) and reflects `mc_iter = 500L` in this
CRAN vignette. Use the canonical run with `mc_iter = 25000L` (chunk
`fit-canonical`, `eval=FALSE`) for publication-quality results. This is a
known discrepancy under investigation (issue \#5: MC STOP calibration audit).
Training ESS and confusion matrix are unaffected.

**Directional analysis.** `direction_map = c("1"=1, "2"=1, "3"=0, "4"=0)`
specifies a fixed categorical partition *a priori*: adjustment types 1 and 2
predict Community motivation; types 3 and 4 predict Individual motivation.
ODA evaluates only this mapping (no partition search) and conducts the MC
permutation test under the same fixed rule  -  a directional hypothesis test
consistent with MPE Chapter 4 categorical/table DIRECTIONAL (Phase 6C).
