---
title: "Getting started with odacore"
author: "odacore"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with odacore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`odacore` is a pure-R implementation of the MegaODA / CTA classification
engine. It provides three main tools:

- **`oda_fit()`** — Optimal Data Analysis (ODA): univariate binary or
  multiclass classification with a single attribute.
- **`cta_fit()`** — Classification Tree Analysis (CTA): recursive ODA-node
  trees with ENUMERATE, LOO STABLE, and MINDENOM endpoint constraints.
- **Translation and graphics** — endpoint staging, propensity weights, tree
  plots, and NOVOmetric bootstrap CIs via `cta_staging_table()`,
  `cta_propensity_weights()`, `novo_boot_ci()`, and `plot.cta_tree()`.

## Binary ODA

`oda_fit()` finds the single cutpoint that maximises Mean PAC (percentage of
accurate classifications, averaged across classes with inverse-frequency
weighting when `priors_on = TRUE`). ESS (Effect Strength for Sensitivity)
measures how far Mean PAC exceeds the chance benchmark of 50%.

```{r oda-binary}
library(odacore)

x <- c(1, 2, 3, 4, 5, 6, 7, 8)
y <- c(0L, 0L, 0L, 0L, 1L, 1L, 1L, 1L)

fit <- oda_fit(x, y, mcarlo = FALSE)
print(fit)
```

The fitted rule, per-class PAC, Mean PAC, and ESS are available directly:

```{r oda-fields}
fit$rule$cut_value    # cutpoint
fit$rule$direction    # which side maps to class 1
fit$ess               # ESS (%)
```

## Classification Tree Analysis

`cta_fit()` grows a binary classification tree in which each split node is an
ODA model. `mindenom` sets the minimum leaf size; `mc_iter` controls the Monte
Carlo permutation test used to screen candidate splits; `loo = "stable"` keeps
only splits whose leave-one-out ESS matches the training ESS.

The example below uses a small synthetic dataset. For publication analyses use
at least `mc_iter = 5000L`.

```{r cta-small}
X <- data.frame(
  x1 = c(1, 2, 3, 4, 5, 6, 7, 8),
  x2 = c(0L, 0L, 1L, 0L, 1L, 1L, 0L, 1L)
)
y <- c(1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L)

tree <- cta_fit(X, y,
  priors_on  = TRUE,
  mindenom   = 2L,
  mc_iter    = 500L,
  mc_seed    = 42L,
  loo        = "off",
  attr_names = c("x1", "x2")
)
print(tree)
```

```{r cta-plot, fig.width = 5, fig.height = 3.5}
plot(tree)
```

## Further reading

The article series (available on the pkgdown site) provides full worked
examples for each component of the package:

| Article | Content |
|---------|---------|
| ODA basics | `oda_fit()` in depth: rule, PAC, ESS, LOO |
| Directional ODA | Chapter 2 (`direction="greater"/"less"`) and Chapter 4 (`direction_map`, `direction="ascending"`) |
| Multiclass ODA | C ≥ 3 class variables; PAC per class; K-class chance benchmark |
| CTA basics | `cta_fit()`: MINDENOM, LOO STABLE, ENUMERATE, pruning |
| Myeloma CTA | MINDENOM family (1 → 30 → 56); canonical computed outputs |
| CTA translation | `cta_staging_table()`, `cta_propensity_weights()`, endpoint reporting |
| MDSA family | `cta_descendant_family()`, `cta_family_table()`, minimum-D selection |
| CTA graphics | `cta_plot_data()` data contract; `plot.cta_tree()` structural and target-enriched diagrams |
| NOVOboot CI | `novo_boot_ci()`: fixed-confusion NOVOmetric bootstrap |
| Validation tiers | CRAN / fast / smoke / full tiers; canon parity; package scope |

The four example-driven CRAN vignettes — *Binary ODA: Migraine Attacks*,
*Binary Ordered Directional ODA: The Refugee Act*, *Binary Categorical
Directional ODA: Gully Erosion*, and *Multiclass Categorical Directional ODA:
Protein Type* — are also available via `browseVignettes("odacore")`.
