---
title: "Multiclass ODA: Convergent Validity of Protein Classification Methods"
author: "odacore"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiclass ODA: Convergent Validity of Protein Classification Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Research question

Nishikawa, Kubota, and Ooi (1983) independently classified 325 proteins into
one of four mutually exclusive types using two different methods: one based on
biological characteristics and one based on amino acid composition.^[Nishikawa,
K., Kubota, Y., & Ooi, T. (1983). Classification of proteins into groups based
on amino acid composition and other characters, II: Grouping into four types.
*Journal of Biochemistry*, 94, 997–1007.] Because the two methods should
theoretically converge on the same type for each protein, the directional
hypothesis is that protein type codes are identical across methods —
demonstrating convergent validity.

Optimal Data Analysis (MultiODA) tests whether amino acid composition type
discriminates biological type, with the *a priori* prediction that type codes
match across methods.

## Data

Biological type (1–4) is the class variable; amino acid composition type (1–4)
is the attribute. Published cell frequencies are reconstructed directly into
observation-level vectors — no external data file is required.

```{r data}
library(odacore)

# Cross-classification: rows = biological type, cols = amino acid type.
# (column-major reconstruction matches published Table 1)
#                  AA=1  AA=2  AA=3  AA=4   total
#  Biological=1     98    16     5     3      122
#  Biological=2     13    50     2     8       73
#  Biological=3      6     4    23    12       45
#  Biological=4      7    19    14    45       85
#  total           124    89    44    68      325

biological_type <- c(
  rep(1L, 98), rep(2L, 13), rep(3L,  6), rep(4L,  7),  # amino_acid = 1
  rep(1L, 16), rep(2L, 50), rep(3L,  4), rep(4L, 19),  # amino_acid = 2
  rep(1L,  5), rep(2L,  2), rep(3L, 23), rep(4L, 14),  # amino_acid = 3
  rep(1L,  3), rep(2L,  8), rep(3L, 12), rep(4L, 45)   # amino_acid = 4
)
amino_acid_type <- c(rep(1L, 124), rep(2L, 89), rep(3L, 44), rep(4L, 68))

table(amino_acid_type, biological_type,
      dnn = c("Amino Acid Type (1-4)", "Biological Type (1-4)"))
```

## Fit the ODA model

Amino acid type is a four-category nominal variable; ODA scans it with the
categorical flag, consistent with the MegaODA reference analysis. Leave-one-out
(LOO) jackknife is not conducted for multicategorical class variables.

```{r fit}
fit <- oda_fit(
  x         = amino_acid_type,
  y         = biological_type,
  attr_type = "categorical",
  mc_iter   = 25000L,
  loo       = "off"
)
```

## Rule and confusion matrix

```{r print-fit}
print(fit)
```

ODA identified the identity mapping across all four types — exactly as
hypothesised *a priori*:

- If amino acid type = 1 → predict biological type = 1
- If amino acid type = 2 → predict biological type = 2
- If amino acid type = 3 → predict biological type = 3
- If amino acid type = 4 → predict biological type = 4

```{r confusion}
# Confusion matrix (actual × predicted); strip dimnames for clean display
conf_mat <- unname(fit$confusion)
rownames(conf_mat) <- paste0("Bio=", 1:4)
colnames(conf_mat) <- paste0("Pred=", 1:4)
print(conf_mat)
```

## ESS / PAC / PV interpretation

```{r metrics}
summary(fit)
```

```{r pac-pv}
m <- oda_metrics(fit)

# PAC (sensitivity) per class
cat("PAC by biological type:\n")
cat("  Type 1:", round(m$pac_by_class[1] * 100, 1), "%\n")
cat("  Type 2:", round(m$pac_by_class[2] * 100, 1), "%\n")
cat("  Type 3:", round(m$pac_by_class[3] * 100, 1), "%\n")
cat("  Type 4:", round(m$pac_by_class[4] * 100, 1), "%\n")

# Predictive value: diagonal / column sums
pv <- diag(fit$confusion) / colSums(fit$confusion) * 100
cat("\nPV by biological type:\n")
cat("  Type 1:", round(pv[1], 1), "%\n")
cat("  Type 2:", round(pv[2], 1), "%\n")
cat("  Type 3:", round(pv[3], 1), "%\n")
cat("  Type 4:", round(pv[4], 1), "%\n")
```

- **PAC (sensitivity per class):** 80.3%, 68.5%, 51.1%, and 52.9% for protein
  types 1 through 4, respectively. Because 25% correct per class is expected by
  chance for a four-class problem, classification of all four types substantially
  exceeds chance.
- **ESS = 50.96%** indicates a relatively strong effect.^[Yarnold, P.R., &
  Soltysik, R.C. (2005). *Optimal Data Analysis: A Guidebook with Software for
  Windows.* Washington, D.C.: APA Books.] All four PAC values exceed 50%,
  confirming that no class is classified below the coin-flip threshold for a
  four-class model (50%).
- **PV:** When the model predicts type 1 it is correct ~79.0% of the time; type
  2, ~56.2%; type 3, ~52.3%; type 4, ~66.2%. All predictive values exceed
  chance.

## Monte Carlo validity

The nondirectional MC p-value is shown in the `summary` output above.

- **MC p-value:** The printed `p(MC)` is a nondirectional Fisher-randomization
  p-value. Interpret by decision threshold (e.g., p < 0.05).
- **LOO:** Leave-one-out jackknife is not conducted for multicategorical class
  variables.

## Notes on reproducibility and current scope

**Fixture parity.** The training rule, confusion matrix, and ESS are verified
against MegaODA.exe output in the package test suite
(`tests/testthat/test-fixture-vignettes.R`, Example 3).

**Directional p-value.** The original analysis specified a *directional*
hypothesis (convergent validity). Directional ODA is not yet implemented
(issue \#6). When available, the directional p-value will be smaller.
