---
title: "Binary ODA: Voting on the Refugee Act of 1980"
author: "odacore"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binary ODA: Voting on the Refugee Act of 1980}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Research question

The Refugee Act of 1980 extended asylum protections in U.S. immigration law.
Because the Act was sponsored by Democrats, a plausible hypothesis is that
support was partisan: Democrats would tend to vote in favor (Pro) and
Republicans against (Con).^[Yarnold, B.M. (1990). *Refugees without refuge:
Formation and failed implementation of U.S. political asylum policy in the
1980's.* Lanham, MD: University Press of America.]

The data below record the vote of 407 U.S. House members alongside their
party affiliation. Optimal Data Analysis (UniODA) is used to determine whether
party affiliation discriminates vote direction, and to quantify the strength of
the association.

## Data

Party affiliation (0 = Republican, 1 = Democrat) is the attribute; vote
(0 = Con, 1 = Pro) is the class variable. Published cell frequencies are
reconstructed directly into observation-level vectors — no external data file
is required.

```{r data}
library(odacore)

# Cross-classification: rows = vote (class), cols = party (attribute).
#          Rep (0)   Dem (1)
#  Con (0)   118       78     n(Con) = 196
#  Pro (1)    34      177     n(Pro) = 211
vote  <- c(rep(0L, 118), rep(0L,  78), rep(1L,  34), rep(1L, 177))
party <- c(rep(0L, 118), rep(1L,  78), rep(0L,  34), rep(1L, 177))

table(vote, party,
      dnn = c("Vote (0=Con, 1=Pro)", "Party (0=Rep, 1=Dem)"))
```

## Fit the ODA model

Party affiliation is binary (0/1); ODA scans it as an ordered attribute
(no categorical flag), which is consistent with the MegaODA reference
analysis. Leave-one-out (LOO) jackknife validity analysis is included.

```{r fit}
fit <- oda_fit(
  x         = party,
  y         = vote,
  attr_type = "ordered",
  mc_iter   = 25000L,
  loo       = "on"
)
```

## Rule and confusion matrix

```{r print-fit}
print(fit)
```

ODA identified a single cut at 0.5, separating the two party values:

- If party ≤ 0.5 (Republican) → predict vote = Con (0)
- If party > 0.5 (Democrat) → predict vote = Pro (1)

This rule is consistent with the directional hypothesis: Democrats
supported the Act and Republicans opposed it.

```{r confusion}
# Confusion matrix: actual vote (rows) × predicted vote (cols)
conf_mat <- matrix(
  c(fit$confusion$TN, fit$confusion$FP,
    fit$confusion$FN, fit$confusion$TP),
  nrow = 2L, byrow = TRUE,
  dimnames = list(Actual    = c("Con(0)", "Pro(1)"),
                  Predicted = c("Con(0)", "Pro(1)"))
)
print(conf_mat)
```

## ESS / PAC / PV interpretation

```{r metrics}
summary(fit)
```

```{r pv}
# Predictive value: accuracy when the model makes a prediction into each class
pv_con <- fit$confusion$TN / (fit$confusion$TN + fit$confusion$FN)
pv_pro <- fit$confusion$TP / (fit$confusion$TP + fit$confusion$FP)
cat("PV Con (0):", round(pv_con * 100, 1), "%\n")
cat("PV Pro (1):", round(pv_pro * 100, 1), "%\n")
```

- **PAC (sensitivity per class):** 60.2% for Republican members (Con vote),
  83.9% for Democratic members (Pro vote). Because 50% correct per class is
  expected by chance, the model classifies Democratic members nearly twice as
  well as chance.
- **ESS = 44.09%** indicates a moderate effect.^[Yarnold, P.R., & Soltysik,
  R.C. (2005). *Optimal Data Analysis: A Guidebook with Software for Windows.*
  Washington, D.C.: APA Books.] The asymmetry (60% vs. 84%) reflects stronger
  partisan signal on the Democratic side — Democrats who sponsored the Act
  voted for it far more uniformly.
- **PV:** When the model predicts a Con vote, it is correct ~77.6% of the time;
  when it predicts a Pro vote, ~69.4%.

## Monte Carlo and LOO validity

The nondirectional MC p-value and LOO result are shown in the `summary` output
above.

- **LOO stability:** The leave-one-out ESS equals the training ESS exactly
  (44.09%), indicating the rule is completely stable — no single observation
  materially alters the model.
- **LOO Fisher exact p < .001:** Statistical significance confirmed in
  holdout analysis.
- **MC p-value:** The printed `p(MC)` is a nondirectional Fisher-randomization
  p-value. Interpret by decision threshold (e.g., p < 0.05), not as an exact
  match to the MegaODA reference value (see Notes).

## Notes on reproducibility and current scope

**Fixture parity.** The training rule, confusion matrix, and ESS are verified
against MegaODA.exe output in the package test suite
(`tests/testthat/test-fixture-vignettes.R`, Example 1).

**MC p-value calibration.** The nondirectional MC p shown here may differ from
MegaODA's reported p = 0.000000 (exact zero). This is a known discrepancy under
investigation (issue \#5: MC STOP calibration audit). Training ESS and
confusion matrix are unaffected.

**Directional p-value.** The original analysis specified a *directional*
hypothesis (one-tailed, a priori). Directional ODA is not yet implemented
(issue \#6). When available, the directional p-value will be smaller.
