---
title: "CTA Tree Diagrams (base-R and ggplot2)"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

<!-- TODO (Slice L artifact pipeline):
  The ggplot2 target-enriched tree example and the myeloma plots depend on
  inst/extdata/articles/myeloma_fits.rds. To unblock:
    1. Build the artifact (see myeloma-cta.Rmd TODO).
    2. Add live ggplot2 tree examples (plot_cta_tree with target_class,
       color_by = "target_rate" vs "prediction") using the myeloma fit.
    3. Update the "v3 renderers" section with full worked examples.

  The synthetic example below is already live and demonstrates the v3 API.
-->

`odacore` provides two levels of CTA tree visualisation:

- **Base-R renderers** (`plot.cta_tree`, `plot.cta_ort`): no external
  dependencies; structural diagram via base graphics.
- **ggplot2 renderers** (`plot_cta_tree`, `plot_lort_tree`): requires the
  ggplot2 package; richer control over colour and layout.

Both share the same data contract: `cta_plot_data()` (for CTA trees) and
`ort_plot_data()` (for LORT trees) produce renderer-independent plot-data
objects. You can pass either the fit object directly or a pre-computed
plot-data object to any renderer.

## `cta_plot_data()`: the renderer-independent data contract

`cta_plot_data()` extracts node/edge geometry, split rules, class counts, and
terminal predictions from a fitted `cta_tree` object. The result is a
self-contained plot-data object usable by any renderer.

```{r synthetic}
library(odacore)

set.seed(1L)
n <- 60L
X <- data.frame(
  x1 = c(rnorm(30, mean = 2), rnorm(30, mean = 5)),
  x2 = c(rnorm(30, mean = 1), rnorm(30, mean = 3))
)
y <- c(rep(1L, 30), rep(2L, 30))

tree <- cta_fit(X, y,
  mindenom   = 12L,
  mc_iter    = 300L,
  mc_seed    = 42L,
  loo        = "off",
  attr_names = c("x1", "x2")
)

pd <- cta_plot_data(tree)
```

## `plot.cta_tree()`: native base-R renderer

```{r base-r-plot, fig.width = 6, fig.height = 4}
plot(tree)
```

## Graphics v3: ggplot2 renderers

`plot_cta_tree()` requires ggplot2 (in Suggests). The function accepts either
the fit object or pre-computed plot-data. `color_by` controls terminal-node fill:

- `"none"` (default): white fill — B/W, publication-ready with no legend.
- `"prediction"`: discrete fill by predicted class.
- `"target_rate"`: continuous gradient by target-class proportion.

```{r ggplot2-tree, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4}
plot_cta_tree(pd, color_by = "prediction")
```

```{r ggplot2-tree-rate, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 6, fig.height = 4}
# target_rate: gradient showing proportion of class 1 per node
plot_cta_tree(pd, color_by = "target_rate", target_class = 1L)
```

**Saving to file:**

```{r ggsave-example, eval = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  p <- plot_cta_tree(pd)
  ggplot2::ggsave("cta_tree.png", p, width = 8, height = 5, dpi = 150)
}
```

## Endpoint label content

Each leaf label shows the predicted class, total n at the endpoint, and
(when `target_class` is set) the raw target-class count and proportion. These
values come directly from the per-leaf `class_counts_raw` stored at fit time
 -  no recomputation occurs inside the renderer.

Endpoint colouring encodes relative target-class proportion *within this tree*
only. It does not imply prevalence, risk, or any external reference standard.

## LORT trees

LORT (Locally Optimal Recursive Trees) are displayed by indexed sub-tree
inspection.  `plot_lort_tree(lort, index = k)` renders the CTA sub-tree at
LORT node `k`.  `show_all = TRUE` returns a named list of ggplot objects.

```{r lort-example, eval = FALSE}
lort <- lort_fit(X, y, mc_iter = 300L, mc_seed = 42L, min_n = 12L)
# Root sub-tree (default index = 1)
plot_lort_tree(lort, color_by = "prediction")
# All sub-trees as a named list
all_plots <- plot_lort_tree(lort, show_all = TRUE)
```

## Myeloma examples

*(To be expanded with artifact pipeline  -  see TODO at top of this file.)*

With the myeloma fit loaded, the full target-enriched tree for MINDENOM=1 and
the stump for MINDENOM=30 will be rendered here with `target_class = 1L` and
`color_by = "target_rate"`.

## Further reading

- `docs/GRAPHICS_V3.md`  -  full v3 function reference, examples, and
  `ggsave` usage
- `articles/cta-basics`  -  CTA fitting and `predict.cta_tree()`
