---
title: "CTA basics with cta_fit()"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Classification Tree Analysis (CTA) grows a binary classification tree in which
every internal node is an ODA model. Each split divides observations into two
branches based on a single attribute cutpoint, and the process recurs until no
further splits pass the MINDENOM, LOO STABLE, and significance criteria. This
article introduces the key concepts and parameters.

## CTA as sequential ODA

CTA answers the question: after the best single ODA rule is applied and
observations are divided into two branches, is there additional discriminating
signal in either branch?

Each node in the tree is an independent ODA fit on the observations routed to
that node. The tree grows top-down, one split at a time. Growth stops at a
node when:

1. The minimum endpoint size `mindenom` is not met by any candidate split.
2. No candidate split passes the LOO STABLE gate (LOO ESS ≈ training ESS).
3. No candidate split reaches significance (MC p < `alpha_split`).

After growing, a backward **pruning** step removes nodes whose sub-tree does
not improve the full-tree ESS relative to the Sidak-adjusted significance
threshold.

## `cta_fit()` arguments

| Argument | Default | Purpose |
|----------|---------|---------|
| `X` | — | Data frame of attribute columns |
| `y` | — | Integer class variable |
| `mindenom` | 10L | Minimum observations at any terminal endpoint |
| `mc_iter` | 5000L | Monte Carlo permutations for significance screening |
| `mc_seed` | 42L | Seed for MC reproducibility |
| `alpha_split` | 0.05 | Significance threshold for accepting a split |
| `loo` | `"stable"` | `"stable"` = LOO STABLE gate; `"off"` = skip LOO |
| `priors_on` | TRUE | Balance class frequencies in objective |
| `attr_names` | NULL | Optional names for attribute columns |

## MINDENOM: controlling endpoint size

`mindenom` sets the minimum number of observations at any terminal leaf. It is
the primary complexity control in CTA: a larger `mindenom` forces broader,
simpler trees; a smaller value allows deeper, more granular trees.

```{r synthetic-data}
library(odacore)

set.seed(1L)
n <- 60L
X <- data.frame(
  x1 = c(rnorm(30, mean = 2), rnorm(30, mean = 5)),
  x2 = c(rnorm(30, mean = 1), rnorm(30, mean = 3))
)
y <- c(rep(1L, 30), rep(2L, 30))
```

```{r fit-mindenom}
# mindenom = 15: at least 15 obs per terminal endpoint
tree <- cta_fit(X, y,
  mindenom    = 15L,
  mc_iter     = 300L,
  mc_seed     = 42L,
  loo         = "off",
  priors_on   = TRUE,
  attr_names  = c("x1", "x2")
)
print(tree)
```

Increasing `mindenom` produces simpler (shallower) trees. Decreasing it allows
more splits but risks overfitting and small-endpoint instability. The
MINDENOM descendant family (see `articles/mdsa-family`) systematically
explores this tradeoff.

## ENUMERATE: root candidate selection

CTA does not simply split on the best attribute and recurse. Instead, it uses
an ENUMERATE phase to evaluate every valid root candidate:

1. **Expanded phase:** for each candidate root attribute, CTA grows the full
   HO-CTA tree below that root and scores the complete tree.
2. **Stump phase:** each root candidate is also scored as a two-leaf stump
   (root split only, path-local scoring).
3. The root with the highest overall ESS (or WESS when weights are active)
   wins.

This ensures the selected root is globally best for the full-tree objective,
not just locally best at the root node.

## LOO STABLE: generalisability gate

`loo = "stable"` (the default) screens candidate splits using the jackknife:

- The full-node model is fit (training ESS).
- For each observation at the node, the model is refit on the remaining n − 1
  cases, and the held-out observation is predicted.
- If |LOO ESS − training ESS| > 1 pp, the split is **UNSTABLE** and rejected.
- Only STABLE splits are eligible for inclusion in the tree.

This prevents the tree from retaining splits that depend critically on
individual observations.

```{r fit-loo}
tree_loo <- cta_fit(X, y,
  mindenom    = 15L,
  mc_iter     = 300L,
  mc_seed     = 42L,
  loo         = "stable",   # default — LOO STABLE gate active
  priors_on   = TRUE,
  attr_names  = c("x1", "x2")
)
print(tree_loo)
```

## Pruning: backward complexity reduction

After the tree is grown, CTA applies a backward pruning step. Starting from
the deepest nodes, each non-terminal sub-tree is removed if it does not
improve the full-tree ESS beyond what a Sidak-adjusted significance threshold
requires. This is analogous to cost-complexity pruning in CART but uses ESS
gain and the Sidak correction for the number of splits evaluated.

The result stored in `$tree` is the **pruned** tree. The pre-pruning
(enumerated) tree is not stored separately.

## `predict.cta_tree()`: path-local missingness

Prediction in CTA is path-local: an observation is routed at each node using
only the attribute at that node. If the attribute is missing for an observation
at the node it is actually traversing, two strategies are available:

```{r predict}
preds_na  <- predict(tree_loo, X, missing_action = "na")       # NA for missing
preds_maj <- predict(tree_loo, X, missing_action = "majority") # legacy majority route

# All non-NA predictions agree:
agree <- sum(preds_na == preds_maj, na.rm = TRUE)
cat("Predictions agree (non-NA):", agree, "of", sum(!is.na(preds_na)), "\n")
```

- `"na"` (canonical): observations with a missing root attribute receive
  `NA_integer_`. They are excluded from scoring.
- `"majority"` (legacy): missing observations are routed to the majority class
  at the node, so all observations receive a prediction.

For scoring and confusion tables the canonical `"na"` mode is used; the
observation count reported by `cta_confusion_table()` excludes unclassified
observations.

## `plot.cta_tree()`: structural diagram

`plot()` dispatches to the base-R tree renderer. It shows the split rule,
node size, and terminal class predictions at each leaf:

```{r plot, fig.width=6, fig.height=4}
plot(tree_loo)
```

For ggplot2 renderers, see `plot_cta_tree()` (requires the ggplot2 package):

```{r ggplot-tree, eval = requireNamespace("ggplot2", quietly = TRUE)}
pd <- cta_plot_data(tree_loo)
plot_cta_tree(pd)
```

## Confusion table

`cta_confusion_table()` extracts the training confusion from the fitted object
without refitting. The confusion is stored as a C × C raw integer matrix at
fit time:

```{r confusion}
cta_confusion_table(tree_loo)
```

## Canonical benchmark: CTA_DEMO

The CTA_DEMO dataset (n = 200, 5 attributes, uniform weights) is the primary
regression fixture for the package. The canonical result with MINDENOM = 1
and `mc_iter = 25000` is:

- Root: V2, cut = 4.5
- Overall ESS = 52.63%
- Attributes V1, V3–V5 are not selected

These values are verified against CTA.exe output in `tests/testthat/test-cta.R`.
See `vignettes/articles/myeloma-cta` for a fully worked MINDENOM family
walkthrough on a real dataset with case weights.

## Further reading

- `articles/myeloma-cta` — full MINDENOM family example (when precomputed
  artifacts are available)
- `articles/cta-translation` — endpoint staging and the translation pipeline
- `articles/mdsa-family` — MINDENOM family and minimum-D model selection
- `articles/cta-graphics` — tree diagram renderers (base-R and ggplot2)
- `docs/CTA_CANON.md` — canonical CTA behavior specification
