---
title: "Directional ODA: Chapter 2 and Chapter 4"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

ODA can incorporate a directional hypothesis stated *a priori*. Specifying a
direction constrains both the training search and the Monte Carlo permutation
test, yielding a one-tailed p-value consistent with the stated hypothesis.
The training ESS and confusion matrix are unaffected  -  only the p-value
changes.

This article covers three directional modes:

| Attribute type | Mode | Argument |
|----------------|------|----------|
| Binary / ordered | Chapter 2 directional | `direction = "greater"` or `"less"` |
| Categorical (fixed partition) | Chapter 4 directional | `direction_map = c(...)` |
| Categorical (identity map, k = C) | Chapter 4 directional | `direction = "ascending"` or `"descending"` |

## Why directional hypotheses?

ODA's nondirectional default searches all possible rules and reports the best
one, using a two-tailed MC p-value. When theory or prior evidence specifies the
direction of the effect *before data collection*, a directional hypothesis is
appropriate:

- It restricts the search (or the MC evaluation) to the hypothesised direction.
- The resulting p-value is one-tailed, which is more powerful under the correct
  direction.
- Using a directional test after seeing the data is not valid  -  the direction
  must be specified *a priori*.

## Chapter 2: binary ordered directional

For a binary or ordered attribute, `direction = "greater"` asserts that larger
attribute values predict class 1; `direction = "less"` asserts that smaller
values predict class 1.

### The Refugee Act example

The Refugee Act of 1980 was sponsored by Democrats. The directional hypothesis
is that Democratic affiliation (party = 1) predicts a Pro vote (vote = 1).^[
Yarnold, B.M. (1990). *Refugees without refuge.* University Press of America.]

```{r refugee-data}
library(oda)

# party: 0 = Republican, 1 = Democrat
# vote:  0 = Con, 1 = Pro
vote  <- c(rep(0L, 118), rep(0L,  78), rep(1L,  34), rep(1L, 177))
party <- c(rep(0L, 118), rep(1L,  78), rep(0L,  34), rep(1L, 177))
```

```{r refugee-fit}
fit_dir <- oda_fit(
  x         = party,
  y         = vote,
  attr_type = "ordered",
  direction = "greater",   # larger party (Democrat) predicts vote = 1 (Pro)
  mc_iter   = 500L,
  mc_seed   = 42L,
  loo       = "on"
)
print(fit_dir)
```

```{r refugee-compare, eval = FALSE}
# For comparison: nondirectional (two-tailed MC p)
fit_nd <- oda_fit(
  x = party, y = vote, attr_type = "ordered",
  mc_iter = 500L, mc_seed = 42L, loo = "on"
)
# ESS is identical; only p(MC) changes (directional p <= nondirectional p / 2)
```

The training rule (party <= 0.5 -> Con; party > 0.5 -> Pro) and ESS = 44.09%
are the same as the nondirectional analysis. The directional p-value is
one-tailed and smaller than the two-tailed p.

**Notes:**

- The attribute here is binary (0/1). For a binary attribute,
  `attr_type = "ordered"` is correct  -  ODA treats it as ordered and finds the
  cut at 0.5, which separates the two values.
- `direction = "greater"` means higher attribute values predict class 1. For a
  binary attribute, this is equivalent to "attribute = 1 predicts class 1."
- LOO is fully supported for binary ordered directional fits.

## Chapter 4: categorical fixed-partition directional

For a nominal attribute, a fixed-partition directional hypothesis assigns each
attribute category to a class *a priori* via `direction_map`. ODA evaluates
only this mapping  -  no partition search is performed. The MC test permutes the
*y* labels while holding the fixed mapping constant.

### The gully erosion example

In a study of gully erosion in southeast Nigeria, four adjustment types were
classified according to whether they require organised community effort.^[Okuh
D, Osumgborogwu IE (2019). *Applied Ecology and Environmental Sciences*, 7,
11-20.] The hypothesis is that Ridges (1) and Shifting Habitation (2) predict
Community motivation, while Relocation (3) and Intensified Cultivation (4)
predict Individual motivation.

```{r gully-data}
motivation  <- c(rep(0L,  85), rep(1L, 173),   # adjustment = 1
                 rep(0L,  65), rep(1L, 170),   # adjustment = 2
                 rep(0L, 172), rep(1L,  10),   # adjustment = 3
                 rep(0L,  45), rep(1L,   0))   # adjustment = 4
adjustment  <- c(rep(1L, 258), rep(2L, 235),
                 rep(3L, 182), rep(4L,  45))
```

```{r gully-fit}
fit_gully <- oda_fit(
  x             = adjustment,
  y             = motivation,
  attr_type     = "categorical",
  direction_map = c("1" = 1L, "2" = 1L, "3" = 0L, "4" = 0L),
  mc_iter       = 500L,
  mc_seed       = 42L,
  loo           = "on"
)
print(fit_gully)
```

Because the prediction rule is fixed *a priori*, every LOO fold applies the
same mapping to the same x values  -  LOO ESS equals training ESS exactly, and
the fit is trivially stable. The MC p-value is directional (one-tailed) because
each permutation evaluates the same fixed partition.

**Notes:**

- `direction_map` is a named integer vector. Names are attribute levels (as
  character), values are class assignments.
- All attribute levels present in `x` must appear in `direction_map`.
- The training ESS and confusion are identical to the nondirectional result
  when the fixed partition happens to be the global optimum  -  which it is here.

**LOO reporting for categorical directional fits:**

With `loo = "on"`, binary fixed-map categorical ODA reports the held-out
confusion, LOO ESS, and a binary one-tailed Fisher LOO p-value. The one-tailed
Fisher LOO p comes from the 2x2 held-out classification table (MPE p. 34); it
is not the MC p-value. MC p and LOO p are separate calculations  -  MC p
permutes *y* labels under the fixed mapping; LOO p comes from the hold-out
table. They are directionally consistent but are not numerically identical.

For multicategorical fits (`direction = "ascending"` or nondirectional with
C > 2 classes), `loo = "on"` reports held-out confusion and ESS but no LOO
p-value. There is no canon-aligned C x C Fisher LOO p for multiclass ODA.

## Chapter 4: categorical identity map (`direction = "ascending"`)

When the attribute has exactly k = C categories matching C class values,
`direction = "ascending"` imposes the identity mapping (attribute value *i*
predicts class *i*) without requiring an explicit `direction_map`.

### The protein type example

```{r protein-data}
biological_type <- c(
  rep(1L, 98), rep(2L, 13), rep(3L,  6), rep(4L,  7),
  rep(1L, 16), rep(2L, 50), rep(3L,  4), rep(4L, 19),
  rep(1L,  5), rep(2L,  2), rep(3L, 23), rep(4L, 14),
  rep(1L,  3), rep(2L,  8), rep(3L, 12), rep(4L, 45)
)
amino_acid_type <- c(rep(1L, 124), rep(2L, 89), rep(3L, 44), rep(4L, 68))
```

```{r protein-fit}
fit_protein <- oda_fit(
  x         = amino_acid_type,
  y         = biological_type,
  attr_type = "categorical",
  direction = "ascending",   # identity map: amino acid i -> biological type i
  mc_iter   = 500L,
  mc_seed   = 42L,
  loo       = "off"
)
summary(fit_protein)
```

With `direction = "ascending"` and k = C = 4, the engine does not search
alternative partitions. The MC test permutes y labels while holding the
identity mapping fixed. Because the identity map is also the globally optimal
categorical mapping for this dataset, training ESS = 50.96% is the same as a
nondirectional analysis; the p-value is smaller (one-tailed).

**Notes:**

- `direction = "ascending"` requires k (number of unique attribute values)
  equals C (number of class values). If they differ, an error is raised.
- `direction = "descending"` reverses the mapping (attribute value k -> class 1,
  attribute value 1 -> class C).
- Multiclass categorical LOO is supported (`loo = "on"`). LOO reports held-out
  confusion and ESS but no Fisher LOO p-value (undefined for C > 2). LOO folds
  search nondirectionally regardless of the `direction` setting; for this dataset
  every fold recovers the identity map, so LOO ESS equals training ESS. This
  example omits LOO for brevity; see `protein-type-multiclass-oda` for the full
  LOO analysis.

## When NOT to use directional: nondirectional default

The nondirectional default (`direction = "both"` for ordered, or no
`direction_map` for categorical) is appropriate when:

- No prior hypothesis specifies the direction.
- The direction is chosen after seeing the data.
- The analysis is exploratory.

Applying a directional constraint that matches the observed direction inflates
the p-value precision without being a valid one-tailed test  -  direction must be
declared *before* the data are seen.

## Further reading

- `vignettes/refugee-act-oda.Rmd`  -  full CRAN vignette for the binary ordered
  directional example
- `vignettes/gully-adjustment-oda.Rmd`  -  full CRAN vignette for the
  fixed-partition categorical directional example
- `vignettes/protein-type-multiclass-oda.Rmd`  -  full CRAN vignette for the
  identity-map multiclass directional example
- `articles/oda-basics`  -  ODA foundations (nondirectional)
- `articles/multiclass-oda`  -  multiclass ODA
