---
title: "MDSA: Descendant Families and Minimum-D Selection"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

<!-- TODO (Slice L artifact pipeline):
  The myeloma family example depends on inst/extdata/articles/myeloma_family.rds.
  To unblock:
    1. Build the artifact (see myeloma-cta.Rmd TODO and build_article_artifacts.R).
    2. Replace the synthetic example below with the myeloma family loaded from
       inst/extdata/articles/myeloma_family.rds.
    3. Verify cta_family_table() values against docs/examples/myeloma-cta-translation.md.

  The synthetic MDSA example below demonstrates the API but uses small n.
-->

MDSA (Model/Descriptor/Scenario Analysis) systematically explores how
classification performance changes as the minimum endpoint size (MINDENOM)
increases. Each MINDENOM value produces a family member — a distinct tree
ranging from a complex multi-split model (small MINDENOM) to a simple stump or
no-tree terminal (large MINDENOM).

## What is MDSA?

Given a fixed dataset, incrementing MINDENOM progressively constrains the
tree structure. The resulting sequence of trees forms a **descendant family**.
Each member represents a valid, self-contained CTA model; the family as a whole
shows the complexity-accuracy tradeoff.

MDSA is used to:

- Identify the simplest model that retains clinically or scientifically
  meaningful classification performance.
- Explore sensitivity to MINDENOM choice.
- Select a model via minimum-D (see below).

## `cta_descendant_family()`: the MINDENOM sequence

`cta_descendant_family()` calls `cta_fit()` at each MINDENOM value in a
user-supplied vector. It returns a `cta_family` object (a list of `cta_tree`
objects plus metadata).

```{r synthetic-family}
library(odacore)

set.seed(1L)
n <- 80L
X <- data.frame(
  x1 = c(rnorm(40, mean = 2), rnorm(40, mean = 6)),
  x2 = c(rnorm(40, mean = 1), rnorm(40, mean = 4))
)
y <- c(rep(1L, 40), rep(2L, 40))

family <- cta_descendant_family(
  X              = X,
  y              = y,
  start_mindenom = 5L,    # first member MINDENOM
  max_steps      = 8L,    # maximum members to produce
  mc_iter        = 300L,
  mc_seed        = 42L,
  loo            = "off",
  attr_names     = c("x1", "x2")
)
print(family)
```

## The D statistic: complexity vs. accuracy

The **D statistic** trades off model complexity against classification
performance. Smaller D is better: it favours simpler models that retain
adequate accuracy.

```{r family-table}
cta_family_table(family)
```

The family table shows D, ESS, the number of terminal endpoints, and the
selected MINDENOM level for each family member.

## `cta_family_table()`: comparing family members

`cta_family_table()` summarises all family members in a single data frame.
Key columns:

| Column | Description |
|--------|-------------|
| `mindenom` | MINDENOM value for this member |
| `strata` | Number of terminal endpoints (NA for no_tree) |
| `overall_ess` | Training ESS (or WESS if weights active) |
| `d` | D statistic |
| `selected_min_d` | TRUE for the family member with minimum D |

## Minimum-D selection

The family member with the smallest D is selected as the **minimum-D model**:
the simplest model consistent with maximum available accuracy. When two members
have equal D, the simpler (larger MINDENOM) member is preferred.

```{r min-d}
ft <- cta_family_table(family)
min_d_row <- ft[which(ft$selected_min_d), ]
cat("Minimum-D member: MINDENOM =", min_d_row$mindenom, "\n")
cat("D statistic:     ", round(min_d_row$d, 4), "\n")
cat("ESS:             ", round(min_d_row$overall_ess, 2), "%\n")
```

## The myeloma family: MINDENOM 1 → 30 → 56

*(To be expanded with artifact pipeline — see TODO at top of this file.)*

The canonical myeloma family demonstrates the full MDSA trajectory:

- **MINDENOM = 1:** V14 → V15 two-split tree (or V17 → V15 after pruning).
  WESS = 27.69% (pruned: 25.43%).
- **MINDENOM = 30:** V17 stump. WESS = 16.51%.
- **MINDENOM = 56:** No admissible splits — leaf only.

All canonical values are in `docs/examples/myeloma-cta-translation.md`.

## Terminal state: no tree

When no MINDENOM candidate produces a valid split (all child sizes < MINDENOM),
the family member is a leaf node — the model predicts the majority class for
all observations. In `cta_family_table()`, this appears as `n_terminals = 1`
and ESS = 0%.

## Further reading

- `articles/myeloma-cta` — full myeloma MINDENOM family walkthrough (when
  precomputed artifacts are available)
- `articles/cta-basics` — CTA fitting fundamentals
- `docs/examples/myeloma-cta-translation.md` — canonical myeloma computed values
