---
title: "Multiclass ODA (C >= 3) with oda_fit()"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

When the class variable has three or more values, `oda_fit()` dispatches to
the MultiODA engine. The objective, chance benchmark, and output fields all
adapt to the multiclass case. This article covers the key differences from
binary ODA.

## The K-class chance benchmark

For a binary problem, random guessing achieves 50% accuracy per class on
average (Mean PAC = 50%, ESS = 0%). For a K-class problem, random guessing
achieves 100/K percent per class on average:

$$\text{chance benchmark} = 100/K \%$$
$$\text{ESS} = \frac{\text{Mean PAC} - (100/K)}{100/K} \times 100\%$$

For K = 3 (three classes), chance is 33.3%; for K = 4, it is 25%. ESS = 0%
still means no better than chance; ESS = 100% still means perfect per-class
classification. The scale is the same regardless of K.

## Ordered multiclass: cut values and segment-class assignments

For an ordered attribute with C classes, the MultiODA engine finds C - 1 cut
values partitioning the attribute range into C contiguous segments, each
assigned to one class. The assignment is monotone when no directional
constraint is given: the engine searches all valid ordered segment-class
assignments and returns the one maximising Mean PAC.

The rule is stored in `$rule$cut_values` (length C - 1) and
`$rule$seg_classes` (length C, one class per segment).

### iris example

```{r iris-3class}
library(oda)

x <- iris$Petal.Length
y <- as.integer(iris$Species)   # 1=setosa, 2=versicolor, 3=virginica

fit3 <- oda_fit(
  x         = x,
  y         = y,
  attr_type = "ordered",
  mcarlo    = FALSE,   # no MC permutation  -  instant; use mc_iter for p-value
  loo       = "off"
)
print(fit3)
```

```{r rule-3class}
fit3$rule$cut_values   # two cutpoints separating three segments
fit3$rule$seg_classes  # class predicted in each segment
```

The three segments divide `Petal.Length` as: setosa (short petals) -> a
boundary -> versicolor (medium) -> a boundary -> virginica (long petals).

## Categorical multiclass: level-class mapping

For a nominal attribute the engine searches all valid partitions of attribute
categories into classes. The best partition maximises Mean PAC. With
`direction = "ascending"` and k = C categories matching C classes, the engine
constrains the search to the identity mapping (category *i* -> class *i*).

### protein type example

Nishikawa et al. (1983) classified 325 proteins into four types by two
independent methods. The convergent-validity hypothesis is that the two
classification systems agree  -  i.e., amino acid type *i* predicts biological
type *i*. `direction = "ascending"` imposes the identity mapping.^[Nishikawa,
K., Kubota, Y., & Ooi, T. (1983). *Journal of Biochemistry*, 94, 997-1007.]

```{r protein-data}
biological_type <- c(
  rep(1L, 98), rep(2L, 13), rep(3L,  6), rep(4L,  7),  # amino_acid = 1
  rep(1L, 16), rep(2L, 50), rep(3L,  4), rep(4L, 19),  # amino_acid = 2
  rep(1L,  5), rep(2L,  2), rep(3L, 23), rep(4L, 14),  # amino_acid = 3
  rep(1L,  3), rep(2L,  8), rep(3L, 12), rep(4L, 45)   # amino_acid = 4
)
amino_acid_type <- c(rep(1L, 124), rep(2L, 89), rep(3L, 44), rep(4L, 68))
```

```{r protein-fit}
fit4 <- oda_fit(
  x         = amino_acid_type,
  y         = biological_type,
  attr_type = "categorical",
  direction = "ascending",    # identity map: AA type i -> biological type i
  mc_iter   = 500L,
  mc_seed   = 42L,
  loo       = "off"
)
print(fit4)
```

## Interpreting per-class PAC at K > 2

```{r pac-k4}
m <- oda_metrics(fit4)
cat("Mean PAC:", round(m$mean_pac * 100, 2), "%  (chance = 25% for K=4)\n")
cat("ESS:     ", round(m$ess, 2), "%\n")
cat("PAC by class:", paste(round(m$pac_by_class * 100, 1), collapse = ", "), "%\n")
```

All four PAC values must exceed the K-class chance benchmark (25%) for the
model to beat chance for every class. PAC below chance for a class indicates
the model is less useful than random prediction for that class  -  even when
Mean PAC and ESS are positive.

**Overall accuracy vs. Mean PAC:**

```{r overall-vs-mean}
# Overall accuracy gives large classes disproportionate influence
overall_acc <- sum(diag(fit4$confusion)) / sum(fit4$confusion) * 100
cat("Overall accuracy:", round(overall_acc, 2), "%\n")
cat("Mean PAC:        ", round(m$mean_pac * 100, 2), "%\n")
```

For balanced class sizes these two measures agree. For imbalanced classes they
diverge, and Mean PAC is the appropriate summary: it treats every class
equally and scales ESS consistently regardless of class frequency.

**Confusion matrix:** For C >= 3, `$confusion` is a C x C integer matrix
(actual x predicted):

```{r confusion-k4}
fit4$confusion
```

## Notes on LOO for multiclass

Leave-one-out is supported for ordered multiclass (`loo = "on"`) and for
unweighted categorical multiclass. Weighted categorical multiclass LOO is
explicitly forbidden  -  the engine raises an error because refit-per-fold
with fixed categorical weights is not well-defined.

```{r loo-3class}
fit3_loo <- oda_fit(
  x         = x,
  y         = y,
  attr_type = "ordered",
  mcarlo    = FALSE,
  loo       = "on"
)
cat("Training ESS:", round(fit3_loo$ess, 2), "%\n")
cat("LOO ESS:     ", round(fit3_loo$loo_ess, 2), "%\n")
cat("LOO status:  ", fit3_loo$loo_status, "\n")
```

## Summary

| Feature | Binary ODA | Multiclass ODA |
|---------|-----------|----------------|
| Class values | 2 | >= 3 |
| Chance benchmark | 50% | 100/K% |
| Ordered rule | 1 cutpoint | C-1 cutpoints |
| Confusion matrix | TN/FP/FN/TP | CxC integer matrix |
| LOO | Supported | Ordered yes; weighted categorical: error |

## Further reading

- `vignettes/protein-type-multiclass-oda.Rmd`  -  full CRAN vignette for the
  four-class protein example
- `articles/directional-oda`  -  Chapter 4 identity-map directional for
  multiclass nominal attributes
- `articles/oda-basics`  -  binary ODA foundations
