---
title: "Fixed-Confusion NOVOmetric Bootstrap CIs"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`novo_boot_ci()` estimates the precision of an observed binary classification
effect by resampling the confusion matrix. The model is held fixed; only
sampling variation in the classification outcome is modelled.

## Fixed-confusion bootstrap: concept and scope

Standard bootstrap confidence intervals for classification performance resample
observations and refit the model on each resample  -  capturing both sampling
variation and model-selection variability. The fixed-confusion NOVOmetric
bootstrap takes a different position: once the model is trained and the
confusion matrix is observed, the question is how precisely that matrix
characterises the true classification performance in the population.

The **fixed-confusion** approach:

1. Expands the C x C confusion matrix to a row-per-observation table of
   (actual, predicted) pairs.
2. On each bootstrap replicate, resamples *k* rows with replacement from that
   table (preserving the observed joint distribution).
3. Computes ESS, sensitivity, specificity, and other metrics for each
   replicate, yielding a **model distribution**.
4. Separately resamples actual and predicted labels *independently* (breaking
   their association) to build a **chance distribution**.
5. Declares novometric significance (Axiom 1) when the model 95% CI lower
   bound exceeds the chance 95% CI upper bound.

What this CI captures:

- Sampling precision of the observed confusion matrix proportions.
- Whether the observed effect could plausibly be a chance result.

What it does not capture:

- Model-selection variability (different data -> different rule).
- Overfitting: if the model was overfit, the confusion matrix used as input
  understates the true error rate.
- Uncertainty in the confusion matrix structure (cells are treated as fixed
  parameters, not as random variables with their own uncertainty).

## `novo_boot_ci()` call structure

```r
novo_boot_ci(
  x,                       # 2x2 integer matrix: actual (rows) x predicted (cols)
  nboot       = 5000L,     # bootstrap replicates
  seed        = NULL,      # integer seed for reproducibility
  sample_frac = 0.5,       # fraction of n sampled per replicate (NOVOboot default)
  probs       = c(0, .025, .05, .25, .5, .75, .95, .975, 1),
  alternative = "two.sided"
)
```

The input `x` must be a 2 x 2 integer matrix constructed with
`byrow = TRUE` using the `[actual, predicted]` convention:

```
         Predicted 0   Predicted 1
Actual 0     TN            FP
Actual 1     FN            TP
```

## Worked example: myeloma MINDENOM=1 confusion

The myeloma MINDENOM=1 pruned tree produces this confusion matrix (verified
against CTA.exe):

```{r conf-matrix}
library(oda)

# Myeloma MINDENOM=1 pruned tree confusion (actual x predicted, byrow = TRUE)
# Rows: actual class (0 and 1). Columns: predicted class (0 and 1).
conf <- matrix(
  c(92, 43,   # actual 0: 92 correct (TN), 43 wrong (FP)
    21, 30),  # actual 1: 21 wrong (FN), 30 correct (TP)
  nrow = 2L, byrow = TRUE,
  dimnames = list(Actual    = c("0", "1"),
                  Predicted = c("0", "1"))
)
print(conf)
```

Note: no myeloma fit is needed  -  the confusion matrix is hardcoded from the
known CTA.exe canonical output.

```{r novoboot}
set.seed(42L)
ci <- novo_boot_ci(conf, nboot = 500L, seed = 42L)
print(ci)
```

## Interpreting the CI: ESS bounds

The `$ci` element provides fixed 95% CI bounds for each metric under both
model and chance distributions:

```{r ci-table}
ci$ci[ci$ci$metric == "ess", ]
```

- **`model_lower`** and **`model_upper`**: 2.5th and 97.5th percentiles of ESS
  under the model bootstrap (resampling the observed confusion).
- **`chance_lower`** and **`chance_upper`**: same, under the chance bootstrap
  (independent resampling of actual and predicted labels).
- **`overlap`**: logical; TRUE if the two 95% CIs overlap.

**Novometric significance (Axiom 1):** `ci$significant == TRUE` when the model
95% CI lower bound exceeds the chance 95% CI upper bound  -  the intervals are
entirely non-overlapping, providing strong evidence that the observed effect
is not a chance result.

```{r significant}
cat("Significant (Axiom 1 CI non-overlap):", ci$significant, "\n")
```

## The `sample_frac` parameter

`sample_frac = 0.5` (the NOVOboot default) samples half the observations per
replicate. This conservatively widens the CIs, making the non-overlap
criterion harder to satisfy and reducing the false-positive rate in small
samples.

For large n, `sample_frac` can be increased toward 1.0 (standard bootstrap).
For publication analyses use `nboot = 5000L` (the default).

## What this CI does not capture

**Model-selection variability:** `novo_boot_ci()` takes the confusion matrix as
fixed input. If the model was chosen based on a data-driven search (as it was
in CTA), the reported confusion matrix reflects in-sample performance. The CI
does not adjust for the fact that a different sample would have selected a
different rule.

**Refit uncertainty:** No refitting occurs. For out-of-sample performance
estimates, use LOO (`loo = "on"` in `oda_fit()` or `cta_fit()`) or held-out
validation before calling `novo_boot_ci()` on the out-of-sample confusion.

**Multi-class extension:** `novo_boot_ci()` operates on 2 x 2 confusion
matrices only (binary class). Multi-class extension is not yet implemented.

## Further reading

- `?novo_boot_ci`  -  complete function documentation with the myeloma example
- `articles/cta-basics`  -  CTA fitting and confusion table
- Yarnold PR (2020). Reformulating the First Axiom of Novometric Theory.
  *Optimal Data Analysis* **9**, 7-8.
- Yarnold PR, Soltysik RC (2016). *Maximizing Predictive Accuracy*. ODA Books.
