---
title: "ODA basics with oda_fit()"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Optimal Data Analysis (ODA) finds the single classification rule  -  a cutpoint
for ordered attributes, or a partition for categorical attributes  -  that
maximises Mean Percentage of Accurate Classifications (Mean PAC) across all
class values. This article walks through the core concepts using a concrete
binary example.

## The ODA problem

Given a class variable **y** (the outcome to classify) and an attribute **x**
(the predictor), ODA asks: what single rule on **x** best separates the classes
in **y**?

For an ordered attribute the rule is a cutpoint *c*:

- observations with x <= *c* -> assigned to one class
- observations with x > *c* -> assigned to the other class

The "best" rule maximises **Mean PAC**: the average, over all classes, of the
per-class accuracy (PAC = percentage of observations in that class correctly
classified). Using the per-class average  -  rather than overall accuracy  - 
ensures that a minority class is not ignored when one class is much larger
than the other.

**ESS (Effect Strength for Sensitivity)** scales Mean PAC against the chance
benchmark of 50%:

$$\text{ESS} = \frac{\text{Mean PAC} - 0.5}{0.5} \times 100\%$$

A model that does no better than random chance has ESS = 0%; perfect
classification has ESS = 100%.

## Fitting with `oda_fit()`

The dataset below reconstructs cell frequencies from Appleton (1995), a
clinical trial in which 67 patients with migraine were randomised to two
treatments and the number of attacks (0-7) recorded.^[Appleton DR (1995).
Pitfalls in the interpretation of studies: III. *Journal of the Royal Society
of Medicine*, 88, 241-243.] The class variable is treatment arm (0 or 1);
the attribute is the ordered attack count.

```{r data}
library(oda)

# Cross-classification: rows = attacks (0-7), cols = treatment arm.
#          T1 (0)  T2 (1)
#  0 att:    13       5
#  1 att:     9      13
#  2 att:     4       6
#  3 att:     2       1
#  4 att:     1       2
#  5 att:     1       3
#  6 att:     3       3
#  7 att:     0       1

treatment <- c(
  rep(0L, 13), rep(1L,  5),   # attacks = 0
  rep(0L,  9), rep(1L, 13),   # attacks = 1
  rep(0L,  4), rep(1L,  6),   # attacks = 2
  rep(0L,  2), rep(1L,  1),   # attacks = 3
  rep(0L,  1), rep(1L,  2),   # attacks = 4
  rep(0L,  1), rep(1L,  3),   # attacks = 5
  rep(0L,  3), rep(1L,  3),   # attacks = 6
  rep(0L,  0), rep(1L,  1)    # attacks = 7
)
attacks <- c(
  rep(0L, 18), rep(1L, 22), rep(2L, 10),
  rep(3L,  3), rep(4L,  3), rep(5L,  4),
  rep(6L,  6), rep(7L,  1)
)
```

`oda_fit()` takes `x` (attribute), `y` (class variable), and an `attr_type`
indicating whether `x` is `"ordered"`, `"binary"`, or `"categorical"`.

```{r fit}
fit <- oda_fit(
  x         = attacks,
  y         = treatment,
  attr_type = "ordered",
  mc_iter   = 500L,    # CRAN-safe; use 25000L for publication
  mc_seed   = 42L,
  loo       = "on"
)
```

## Reading the output: rule, PAC, ESS

```{r print-fit}
print(fit)
```

ODA identified a cut at 0.5 (zero attacks vs. one or more). The `$rule`
element records the classification rule:

```{r rule}
fit$rule$cut_value   # cutpoint
fit$rule$direction   # which side maps to which class
```

`"right_class_1"` means observations with attacks > 0.5 are assigned to
class 1 (Treatment 2); observations <= 0.5 (zero attacks) are assigned to
class 0 (Treatment 1).

The full `summary()` displays per-class PAC, Mean PAC, ESS, and the MC
p-value:

```{r summary}
summary(fit)
```

**Interpreting PAC and ESS:**

- PAC for class 0 (T1) = percentage of T1 patients correctly classified.
- PAC for class 1 (T2) = percentage of T2 patients correctly classified.
- Mean PAC = their unweighted average.
- ESS = (Mean PAC - 50%) / 50% x 100%. Here ESS ~ 24.69%.

**Confusion matrix:** The raw integer confusion is in `$confusion` (binary
form: TN, FP, FN, TP):

```{r confusion}
cat("TN =", fit$confusion$TN, " FP =", fit$confusion$FP, "\n")
cat("FN =", fit$confusion$FN, " TP =", fit$confusion$TP, "\n")
```

**Predictive value (PV):** When the model predicts a class, how often is it
correct?

```{r pv}
pv_t1 <- fit$confusion$TN / (fit$confusion$TN + fit$confusion$FN)
pv_t2 <- fit$confusion$TP / (fit$confusion$TP + fit$confusion$FP)
cat("PV T1 (0):", round(pv_t1 * 100, 1), "%\n")
cat("PV T2 (1):", round(pv_t2 * 100, 1), "%\n")
```

## Leave-one-out (LOO) validity

`loo = "on"` runs a true jackknife: each observation is dropped in turn, the
model is refit on the remaining n - 1 cases, and the held-out observation is
predicted by the refitted rule.

```{r loo}
cat("Training ESS:", round(fit$ess,          2), "%\n")
cat("LOO ESS:     ", round(fit$loo$ess_loo,  2), "%\n")
loo_stable <- isTRUE(all.equal(fit$ess, fit$loo$ess_loo, tolerance = 1e-4))
cat("LOO status:  ", if (loo_stable) "STABLE" else "UNSTABLE", "\n")
```

`LOO status: STABLE` means the LOO ESS matches the training ESS (within a
small numerical tolerance). Complete LOO stability means no single observation
drives the rule  -  the model generalises.

For binary ordered attributes where the cut falls at a sparse boundary (e.g.
attacks = 0 here), LOO is often trivially stable: every fold produces the same
cut.

## Notes on MC p-value calibration

The Monte Carlo p-value tests whether the observed ESS could arise by chance.
ODA's chance benchmark is 50% Mean PAC (ESS = 0%), not overall accuracy.

The `p(MC)` shown reflects `mc_iter = 500L`. For publication, use
`mc_iter = 25000L` (the canonical MegaODA reference run). The training ESS
and confusion matrix are unaffected by `mc_iter`; only the p-value precision
changes with low iteration counts.

## Accessor functions

`oda_predictions()`, `oda_confusion()`, and `oda_metrics()` provide
programmatic access to fit components:

```{r accessors}
m <- oda_metrics(fit)
cat("Mean PAC:", round(m$mean_pac * 100, 2), "%\n")
cat("ESS:     ", round(m$ess, 2), "%\n")
```

## Key ESS thresholds

| ESS range | Conventional interpretation     |
|-----------|----------------------------------|
| < 25%     | Relatively weak effect           |
| 25-50%    | Moderate effect                  |
| 50-75%    | Relatively strong effect         |
| > 75%     | Strong effect                    |

The migraine example (ESS ~ 24.69%) is marginally below the moderate
threshold  -  a weak but non-chance association.^[Yarnold, P.R., & Soltysik,
R.C. (2005). *Optimal Data Analysis: A Guidebook with Software for Windows.*
Washington, D.C.: APA Books.]

## Further reading

- `vignettes/migraine-attacks-oda.Rmd`  -  full CRAN vignette for this example
- `articles/directional-oda`  -  Chapter 2 and Chapter 4 directional constraints
- `articles/multiclass-oda`  -  extending to C >= 3 classes
- `articles/cta-basics`  -  Classification Tree Analysis
