---
title: "oda practitioner guide"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

This guide walks through the full oda workflow in order: power planning,
data preparation, ODA fitting, CTA and recursive trees, Structural
Decomposition Analysis, covariate balance diagnostics, graphics, and bootstrap
confidence intervals.  All examples use public, exported functions only.  Code
chunks are self-contained and reproducible.

**Scope note.** oda implements unit-weighted binary and multiclass ODA,
Classification Tree Analysis (CTA), Locally Optimal Recursive Trees (LORT),
Structural Decomposition Analysis (SDA), and associated reporting and diagnostic
tools. Weighted SDA, multiclass CTA, SORT, and GORT are deferred and not yet
implemented.  See the final section for a complete boundary list.

**Relationship to the legacy ODA package.** The legacy `ODA` R package
provides an R interface for MegaODA.exe and CTA.exe workflows and their outputs.
`oda` is a separate native-R implementation of core ODA, CTA, and selected
novometric workflows, with test-backed public objects, evidence accessors,
validation helpers, graphics, and practitioner examples.  `oda` is a native-R
core continuation of the ODA software ecosystem, not a drop-in replacement for
every MegaODA/CTA.exe feature.  Use `ODA` for historical executable-interface
workflows.  Use `oda` for native R workflows that are implemented, documented,
and tested in this package.

---

## 1. Installation

Install from GitHub (development version) or from CRAN once published:

```{r install, eval = FALSE}
# Development version
remotes::install_github("njrhodes/oda_r")

# CRAN (once released)
# install.packages("oda")

library(oda)
```

```{r load, include = FALSE}
library(oda)
```

---

## 2. Data readiness

`oda_readiness_check()` audits a predictor frame and class vector for common
problems: wrong types, missing values in coded form, class imbalance, and
empty columns.  Run it before any ODA or CTA fit.

```{r readiness-data}
set.seed(2024L)
n     <- 100L
y     <- rep(c(0L, 1L), each = 50L)
x1    <- c(round(rnorm(50L, mean = 40, sd = 10)),
           round(rnorm(50L, mean = 56, sd = 10)))
x2    <- c(round(rnorm(50L, mean = 24, sd =  9)),
           round(rnorm(50L, mean = 30, sd =  9)))
x3    <- as.integer(c(rbinom(50L, 1L, 0.25), rbinom(50L, 1L, 0.65)))
treat <- as.integer(rep(c(0L, 1L), times = 50L))
dat   <- data.frame(y = y, x1 = x1, x2 = x2, x3 = x3, treat = treat)

X <- dat[, c("x1", "x2", "x3")]
```

```{r readiness-check}
rdy <- oda_readiness_check(X = X, y = dat$y)
rdy$ok
rdy$attr_types
```

`oda_clean_missing_codes()` replaces a numeric missing code (e.g., -9) with
`NA` across a data frame:

```{r clean-missing, eval = FALSE}
# Replace -9 with NA in all columns
X_clean <- oda_clean_missing_codes(X, codes = -9)
```

`oda_infer_attr_types()` returns a named character vector of inferred ODA
attribute types for a predictor frame:

```{r infer-types}
oda_infer_attr_types(X)
```

---

## 3. Power and sample-size planning

**Scope: unit-weighted binary 2-class, binary (2-level) attribute only.**
This is the Fisher exact / ODA isomorphism planning case described by
Rhodes (2020) and anchored in Yarnold and Soltysik (2005).  It is not a CTA,
LORT, SDA, weighted, or multiclass power method.

### 3.1 `oda_power()`: Monte Carlo power estimation

Supply either per-group proportions (`p1`, `p2`) or an ESS target.  ESS maps
to proportions under the symmetric balanced convention:
`p2 = ESS/200 + 0.5`, `p1 = 1 - p2`.

```{r power-basic}
# Power at n = 50 per group for ESS = 48%
pw <- oda_power(n1 = 50L, ess = 48, nsim = 2000L, mc_seed = 42L)
print(pw)
```

```{r power-curve}
# Power curve: vary n, fixed ESS
pw_curve <- oda_power(
  n1      = c(30L, 50L, 80L, 120L),
  ess     = 48,
  nsim    = 2000L,
  mc_seed = 42L
)
print(pw_curve)
```

```{r power-direct}
# Direct proportion input (asymmetric design)
pw_direct <- oda_power(
  n1 = 40L, n2 = 60L,
  p1 = 0.30, p2 = 0.65,
  nsim    = 2000L,
  mc_seed = 42L
)
print(pw_direct)
```

Sidak correction for multiple comparisons:

```{r power-sidak}
pw_sid <- oda_power(
  n1      = 80L,
  ess     = 48,
  comp    = 3L,       # 3 comparisons; Sidak-adjusts alpha
  nsim    = 2000L,
  mc_seed = 42L
)
pw_sid$alpha_adj    # adjusted alpha
pw_sid$power
```

### 3.2 `oda_sample_size()`: minimum n for target power

Bisects over n to find the smallest balanced sample size achieving
`power_target` (default 0.80).  Use `nsim` >= 5000L and a fixed `mc_seed`
for published results; the example uses lower `nsim` for speed.

```{r sample-size}
ss <- oda_sample_size(
  power_target = 0.80,
  ess          = 48,
  nsim         = 2000L,
  mc_seed      = 42L
)
print(ss)
```

---

## 4. Univariate ODA (`oda_fit()`, binary class)

ODA finds the single rule on one attribute that maximises Mean PAC across all
class values.  For an ordered attribute the rule is a cutpoint; for a
categorical attribute it is a partition; for a binary attribute it is a
direct class assignment.

**Effect size.** ESS (Effect Strength for Sensitivity) scales Mean PAC against
the 50% chance baseline:

```
ESS = (Mean PAC - 0.50) / 0.50 * 100%
```

ESS = 0% is chance; ESS = 100% is perfect classification.

```{r oda-fit}
fit1 <- oda_fit(
  x         = dat$x1,
  y         = dat$y,
  attr_type = "ordered",
  mc_iter   = 2000L,
  mc_seed   = 42L
)
print(fit1)
```

```{r oda-summary}
summary(fit1)
```

```{r oda-accessors}
oda_metrics(fit1)      # ESS, Mean PAC, p-value
oda_d_stat(fit1)       # D-statistic (classification complexity)
oda_confusion(fit1)    # raw confusion matrix
```

Apply the fitted rule to new data with `predict()`:

```{r oda-predict}
preds <- predict(fit1, newdata = dat$x1)
table(predicted = preds, actual = dat$y)
```

### 4.1 Binary attribute

```{r oda-binary}
fit_bin <- oda_fit(
  x         = dat$x3,
  y         = dat$y,
  attr_type = "binary",
  mc_iter   = 2000L,
  mc_seed   = 42L
)
print(fit_bin)
```

### 4.2 Missing-code handling

Pass numeric missing codes to `miss_codes`; those values are excluded from
the fit and scored as `NA` in predictions:

```{r oda-miss, eval = FALSE}
fit_miss <- oda_fit(
  x          = dat$x1,
  y          = dat$y,
  miss_codes = c(-9, -99),
  mc_iter    = 2000L,
  mc_seed    = 42L
)
```

### 4.3 Leave-one-out (LOO) gate

`loo = "stable"` applies the MegaODA LOO STABLE criterion: a rule is accepted
only when the LOO ESS approximates the training ESS (|delta| <= 0.01 pp).
Use `loo = "off"` (default) when computing power-conservative estimates; use
`"stable"` for final confirmatory analyses.

```{r oda-loo}
fit_loo <- oda_fit(
  x         = dat$x1,
  y         = dat$y,
  attr_type = "ordered",
  loo       = "stable",
  mc_iter   = 2000L,
  mc_seed   = 42L
)
oda_metrics(fit_loo)
```

---

## 5. Multiclass ODA (`oda_fit()`, three or more classes)

When `y` has three or more distinct values, `oda_fit()` dispatches to the
multiclass engine.  The rule partitions the attribute into C ordered segments
(for ordered attributes) or assigns categories to classes (for categorical
attributes).

```{r multiclass-fit}
# Iris: predict species (3 classes) from petal length
petal_fit <- oda_fit(
  x         = iris$Petal.Length,
  y         = as.integer(iris$Species),
  attr_type = "ordered",
  mc_iter   = 2000L,
  mc_seed   = 42L
)
print(petal_fit)
```

```{r multiclass-summary}
summary(petal_fit)
oda_metrics(petal_fit)
```

```{r multiclass-predict}
petal_preds <- predict(petal_fit, newdata = iris$Petal.Length)
table(predicted = petal_preds, actual = as.integer(iris$Species))
```

Multiclass categorical example -- predict iris species from the species column
re-encoded as a factor (illustrative only):

```{r multiclass-categorical}
petal_cat <- oda_fit(
  x         = iris$Petal.Width,
  y         = as.integer(iris$Species),
  attr_type = "ordered",
  mc_iter   = 2000L,
  mc_seed   = 42L
)
oda_metrics(petal_cat)
```

---

## 6. Classification Tree Analysis (`cta_fit()`)

CTA grows a binary tree in which every internal node is an ODA model.  Each
split divides the working sample into two branches on a single attribute
cutpoint and the process recurses until no split passes the MINDENOM, LOO
STABLE, and significance criteria.  A backward pruning step then removes nodes
whose sub-tree does not improve full-tree ESS.

**Key arguments:**

| Argument | Default | Role |
|----------|---------|------|
| `mindenom` | 10L | Minimum terminal endpoint size |
| `mc_iter` | 5000L | Monte Carlo permutations for significance |
| `mc_seed` | 42L | RNG seed |
| `alpha_split` | 0.05 | Significance threshold for a split |
| `loo` | `"stable"` | LOO STABLE gate (`"off"` skips) |
| `priors_on` | TRUE | Balance classes in objective |

```{r cta-fit}
tree <- cta_fit(
  X           = X,
  y           = dat$y,
  mindenom    = 10L,
  mc_iter     = 1000L,
  mc_seed     = 42L,
  loo         = "stable",
  priors_on   = TRUE,
  attr_names  = c("x1", "x2", "x3")
)
print(tree)
```

```{r cta-node-table}
cta_node_table(tree)
```

```{r cta-confusion}
cta_confusion_table(tree)
```

```{r cta-d}
cta_d_stat(tree)     # D = number of terminal endpoints - 1
```

```{r cta-predict}
preds_cta <- predict(tree, newdata = X)
table(predicted = preds_cta, actual = dat$y)
```

### 6.1 MINDENOM family

The MINDENOM family is the sequence of trees obtained by increasing MINDENOM
from 1 to n.  The min-D member is the simplest tree that still achieves the
same classificatory performance as the most complex member.

```{r cta-family}
fam <- cta_descendant_family(
  X          = X,
  y          = dat$y,
  mc_iter    = 300L,
  mc_seed    = 42L,
  max_steps  = 5L    # increase for production
)
cta_family_table(fam)
```

---

## 7. CTA reporting and propensity

All reporting functions compute on demand from the stored tree object.
Nothing extra is stored at fit time (lean-fit invariant).

```{r cta-staging}
# Endpoints ordered by ascending target-class proportion (lowest to highest risk)
cta_staging_table(tree)
```

```{r cta-endpoint-summary}
cta_endpoint_summary(tree)
```

```{r cta-propensity}
# Stabilized propensity-style weights per endpoint x class
cta_propensity_weights(tree)
```

```{r cta-assign}
# Assign training observations to terminal endpoints (on demand)
ep <- cta_assign_endpoints(tree, newdata = X)
head(ep)
```

```{r cta-obs-weights}
# Map each observation to its endpoint-level weight
ow <- cta_observation_weights(tree, newdata = X, y = dat$y)
head(ow)
```

---

## 8. Locally Optimal Recursive Trees (LORT)

LORT (`lort_fit()`) applies the greedy local min-D strategy: at each terminal
endpoint it scans the MINDENOM descendant family, selects the min-D member,
and recurses until no further structure is found or a compute guard fires.

**Scope:** LORT is adjacent workflow-layer composition using canonical CTA and
MDSA components.  It is not Chapter 12 SORT canon, does not use a global search,
and does not require an SDA source object.

```{r lort-fit}
lort <- lort_fit(
  X                = X,
  y                = dat$y,
  mc_iter          = 200L,   # increase to 5000L for production
  mc_seed          = 42L,
  min_n            = 20L,
  max_depth        = 4L,     # safety cap for illustration
  family_max_steps = 5L,     # limit MDSA depth per node for illustration
  loo              = "off",  # "stable" recommended for production
  verbose          = FALSE
)
print(lort)
```

```{r lort-strata}
cta_strata(lort)         # one row per leaf
```

```{r lort-predict}
lort_preds <- predict(lort, newdata = X)
table(predicted = lort_preds, actual = dat$y)
```

```{r lort-node-table}
cta_ort_node_table(lort)
```

NOVOboot works directly on a `cta_ort` object (see Section 12).

---

## 9. Structural Decomposition Analysis (`sda_fit()`)

SDA fits a sequence of ODA models.  After each step the correctly classified
observations are removed and fitting continues on the residual sample.  The
procedure terminates when no attribute reaches significance on the residual, or
when a safety guard fires.

**Scope:** SDA is a procedure that traverses the attribute space by class; it
is not a model.  Weighted SDA is deferred (SDA-5).  SDA does not estimate
propensity scores.

Two modes:
- `"unioda_max_ess"`: iterative UniODA, selects the attribute with the highest
  ESS at each step.
- `"novometric_min_d"`: per-attribute MDSA via `cta_descendant_family()`;
  requires `mindenom`.

```{r sda-fit}
sda <- sda_fit(
  X       = X,
  y       = dat$y,
  mode    = "unioda_max_ess",
  mc_iter = 1000L,
  mc_seed = 42L,
  verbose = FALSE
)
print(sda)
```

```{r sda-steps}
sda_step_table(sda)
```

```{r sda-selected}
sda_selected_attributes(sda)
```

```{r sda-candidate}
sda_candidate_table(sda)
```

### 9.1 SDA planning with `auto_sda_plan()`

`auto_sda_plan()` is a pre-analysis dry-run planner.  It audits a data frame
for SDA readiness and reports the candidate set, excluded columns, class
balance, and mode.  It does not fit any models.

```{r sda-plan}
# Pass the full data frame and the name of the outcome column
plan <- auto_sda_plan(
  data    = dat,
  outcome = "y",
  mode    = "unioda_max_ess",
  dry_run = TRUE
)
print(plan)
```

### 9.2 SDA to CTA bridge

After SDA identifies the discriminating attribute sequence, use
`sda_to_cta_data()` to prepare data for a downstream CTA fit restricted to
the SDA-selected attributes:

```{r sda-cta-bridge}
cta_data <- sda_to_cta_data(sda, X = X, y = dat$y)
# Returns list(X_cta, y_cta); X_cta contains only SDA-selected columns
head(cta_data$X_cta)
```

```{r as-cta-candidates}
candidates <- as_cta_candidates(sda)
print(candidates)
```

---

## 10. Covariate balance diagnostics

`oda_balance_table()` runs a univariate ODA model for each covariate with
`group` as the class variable.  The result quantifies how well each covariate
discriminates between groups before and, optionally, after weighting.

```{r balance-fit}
bal <- oda_balance_table(
  group   = dat$treat,
  X       = dat[, c("x1", "x2", "x3")],
  mc_iter = 1000L,
  mc_seed = 42L,
  adjust  = "none"
)
print(bal)
```

```{r balance-smd}
# SMD companion table (standardised mean differences)
smd <- smd_balance_table(
  group = dat$treat,
  X     = dat[, c("x1", "x2", "x3")]
)
print(smd)
```

CTA multivariate balance table -- fits its own CTA with `group` as the class
variable to characterise joint covariate imbalance across endpoints:

```{r cta-balance}
cta_bal <- cta_balance_table(
  group   = dat$treat,
  X       = dat[, c("x1", "x2", "x3")],
  mc_iter = 500L,
  mc_seed = 42L
)
print(cta_bal)
```

---

## 11. Graphics

Graphics require ggplot2 (listed in Suggests).  All renderer functions are pure
renderers: they never call fitting functions.  Pass the fitted object or
pre-computed plot-data directly.

```{r has-ggplot2, include = FALSE}
has_gg <- requireNamespace("ggplot2", quietly = TRUE)
```

### 11.1 CTA tree diagram

Default rendering is B/W with no color or legend — suitable for direct journal
submission.  Pass `color_by = "target_rate"` or `color_by = "prediction"` for
colored endpoint fills.

```{r cta-tree-plot, eval = has_gg, fig.width = 7, fig.height = 5}
# Default: B/W, no ESS on nodes, simple labels
plot_cta_tree(tree, target_class = 1L)
```

The data contract layer (`cta_plot_data()`) is renderer-independent and can be
inspected or passed to custom renderers:

```{r cta-plot-data}
pd <- cta_plot_data(tree)
names(pd)
```

### 11.2 LORT sub-tree inspection

`plot_lort_tree()` renders a single LORT node's CTA sub-tree (indexed inspection).
Use `index` to pick a specific node; `show_all = TRUE` returns a named list of
ggplot objects.

```{r lort-plot, eval = has_gg, fig.width = 7, fig.height = 5}
# Root sub-tree (node 1)
plot_lort_tree(lort, index = 1L, target_class = 1L)
```

### 11.2b CTA family member

`plot_cta_family()` renders a single member of a MDSA MINDENOM family.  Use
`index = "min_d"` to plot the member with minimum D-statistic (`fam` was fitted
in Section 6).

```{r family-plot, eval = has_gg, fig.width = 7, fig.height = 5}
plot_cta_family(fam, index = "min_d", target_class = 1L)
```

### 11.3 ODA covariate balance plot

```{r balance-plot, eval = has_gg, fig.width = 6, fig.height = 4}
plot_oda_balance(bal)
```

### 11.4 SMD balance and Love plot

```{r smd-plot, eval = has_gg, fig.width = 6, fig.height = 4}
plot_smd_balance(smd)
```

```{r love-plot, eval = has_gg, fig.width = 6, fig.height = 4}
plot_balance_love(smd)
```

### 11.5 CTA balance plot

```{r cta-balance-plot, eval = has_gg, fig.width = 6, fig.height = 4}
plot_cta_balance(cta_bal)
```

---

## 12. Novometric bootstrap (`novo_boot_ci()`)

NOVOboot estimates the precision of an observed classification effect by
comparing model and chance distributions via permutation bootstrap.  The
method is based on Yarnold (2020) and Yarnold and Soltysik (2016).

**Fixed-confusion bootstrap:** `novo_boot_ci()` samples from the stored
confusion matrix.  It does not refit ODA or CTA models and does not estimate
model-selection variability.

Novometric significance (Axiom 1) is declared when the 95% confidence
intervals for model ESS and chance ESS do not overlap.

### 12.1 From an ODA fit

```{r novoboot-oda}
ci_oda <- novo_boot_ci(fit1, nboot = 2000L, seed = 42L)
print(ci_oda)
```

### 12.2 From a CTA tree

```{r novoboot-cta}
ci_cta <- novo_boot_ci(tree, nboot = 2000L, seed = 42L)
print(ci_cta)
```

Node-level bootstrap (single terminal endpoint):

```{r novoboot-cta-node}
ep_ids <- cta_strata(tree)$endpoint_node_id
if (length(ep_ids) >= 1L) {
  ci_node <- novo_boot_ci(tree, node_id = ep_ids[1L],
                          nboot = 1000L, seed = 42L)
  print(ci_node)
}
```

### 12.3 From a LORT tree

```{r novoboot-lort}
ci_lort <- novo_boot_ci(lort, nboot = 2000L, seed = 42L)
print(ci_lort)
```

### 12.4 From a raw 2x2 matrix

```{r novoboot-matrix}
m <- matrix(c(40L, 10L, 8L, 42L), nrow = 2L,
            dimnames = list(actual    = c("0", "1"),
                            predicted = c("0", "1")))
ci_mat <- novo_boot_ci(m, nboot = 2000L, seed = 42L)
print(ci_mat)
```

`as_confusion_matrix()` converts a tidy confusion data frame (e.g., from
`cta_confusion_table()`) to the 2x2 matrix form required by the default
method:

```{r as-confusion-matrix}
ct  <- cta_confusion_table(tree)
m2  <- as_confusion_matrix(ct)
print(m2)
```

---

## 13. Production checklist

Before committing to a final analysis, verify each item:

**Data preparation**

1. Run `oda_readiness_check()`.  Confirm `rdy$ok == TRUE`; resolve all issues
   before proceeding.
2. Apply `oda_clean_missing_codes()` if numeric missing codes are present.
   Verify the count of `NA` values introduced matches the expected missing count.
3. Confirm attribute types via `oda_infer_attr_types()`.  Override with
   `attr_types` argument when auto-detection is ambiguous.

**ODA**

4. Use `mc_iter >= 25000L` (the MegaODA default) for published significance
   tests.  Lower values (e.g., 1000-5000) are acceptable for exploratory work.
5. Use `loo = "stable"` for confirmatory analyses.  Verify `loo_status` in the
   fit summary is `"STABLE"` for the accepted rule.
6. Report ESS (not overall accuracy) as the primary effect measure.

**CTA**

7. Select `mindenom` based on the minimum acceptable endpoint size, not by
   tuning on model performance.  Document the chosen value and its rationale.
8. Confirm `cta_node_table()` shows the expected tree structure.
9. Use `cta_staging_table()` to verify endpoint risk ordering before reporting.
10. Use `cta_propensity_weights()` for endpoint-level stabilized weights; use
    `cta_assign_endpoints()` + `cta_observation_weights()` when observation-
    level weights are needed.

**Effect size and bootstrap**

11. Report D-statistic (`cta_d_stat()`) alongside ESS; it quantifies
    classification complexity.
12. Run `novo_boot_ci()` with `nboot >= 5000L` for final confidence intervals.
    Confirm Axiom 1 significance (non-overlapping 95% CIs for model vs. chance).

**Power planning (prospective studies)**

13. Use `oda_power()` with `nsim >= 10000L` and a fixed `mc_seed` for
    published power estimates.
14. Use `oda_sample_size()` with `mc_seed` fixed so the bisection result is
    reproducible across sessions.

---

## 14. What oda does not yet do

The following capabilities are explicitly deferred or reserved:

| Capability | Status |
|-----------|--------|
| SORT (`sort_fit()`) | Reserved -- requires SDA anchor; not implemented |
| GORT (`gort_fit()`) | Reserved -- future design only; not implemented |
| Weighted SDA | Deferred (SDA-5); weighted SDA will error if attempted |
| SDA-derived propensity weights | Deferred; SDA produces step order, not propensity strata |
| Multiclass CTA | Not yet scoped |
| Multiclass / weighted / CTA power | Not yet scoped |
| Generic propensity API | Not planned |

If a capability you need is on this list, it is a design gap, not a usage
error.  Check `docs/STATUS.md` in the package source for the current
implementation state.

---

## References

Rhodes, N. J. (2020). Statistical power analysis in ODA, CTA and Novometrics.
*Optimal Data Analysis*, 9. https://odajournal.files.wordpress.com/2020/02/v9a5.pdf

Yarnold PR, Soltysik RC (2005). *Optimal Data Analysis: A Guidebook with
Software for Windows.* Washington, DC: APA Books.

Yarnold PR, Soltysik RC (2016). *Optimal Data Analysis: A Guidebook with
Software for Windows (2nd ed.)* Washington, DC: APA Books.

Yarnold PR, Linden A (2017). Novometric analysis of cross-validated
ODA models. *Optimal Data Analysis*, 6, 30-39.

Yarnold PR (2020). *Maximizing Predictive Accuracy.* Chicago, IL:
ODA Books.
